---
title: "Getting Started: the OzoneExposure package"
author: "Matt Espe"
date: "2 Apr 2019"
output: pdf_document
vignette: 
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{rmarkdown::render}
  \usepackage[utf8]{inputenc}
---

# Introduction to the package

The OzoneExposure package was created to facilitate the fitting and
comparison of two model for the effects of ozone exposure in
humans. These two models are from the following publications:

+ William F. McDonnell, Paul W. Stewart, Marjo V. Smith, Chong S. Kim
  & Edward S. Schelegle (2012) Prediction of lung function response
  for populations exposed to a wide range of ozone conditions,
  Inhalation Toxicology, 24:10, 619-633, DOI:
  10.3109/08958378.2012.705919

+ Edward S. Schelegle, William C. Adams, William F. Walby & M. Susan
  Marion (2012) Modelling of individual subject ozone exposure response
  kinetics, Inhalation Toxicology, 24:7, 401-415
  
Both models predict changes in the FEV1 measure (delta FEV1 or dFEV1
hereafter) over time after exposure to different ozone
concentrations. The data used to fit these models is from a series of
experiments conducted at UC-Davis and by the EPA. Each model has been
coded in Stan, a language for probabilistic inference (see
[mc-stan.org](mc-stan.org)), to increase the speed at which
calculations are done and hence to speed up the model fitting
process. Each model is compiled on package install. 

## Model fitting procedures

We discovered that both models can confound a traditional numerical
optimizer due to multiple local maxima present in the response
surface. Therefore, each model needs a slightly different approach in
order to find the global optima. 

For the Schelegle model: This model only has 4 parameters when fit
using the log-likelihood: DOS, K, A, and the residual error
(simga). Therefore, we found that a grid search was possible to
explore the response surface. In short, we form a grid of possible
parameter combinations, then run the model with these parameters set
and calculate the fit statistic (e.g., the AIC) for each
combination. Since the model code is compiled for compuational speed,
each iteration of this takes roughly 0.001 seconds on a reasonably
fast laptop. 

For the McDonnell model: Since this model utilizes a random effect
parameter for each individual, as well as 8-10 unique fixed
parameters, the number of possible combinations of parameters quickly
escilates to the point where a grid search is not possible. Therefore,
the model is fit repeatedly from different random points, with the
optima for each random start recorded. Given a high number of random
starting points, different optima can be found and compared to
identify the one that gives the best model fit.

# Using the package

## Quick start

### Reading data

These models have been coded in a way that they can be fit to the
exact same input data. This has been done to facilitate comparisons
between models when subjects are removed from the dataset.

First, the data have been archived as SAS formated files with the
following names:

+ z9312_epa.sas7bdat
+ z9312_kim.sas7bdat
+ z9312_ucd.sas7bdat

To read these into R, we first provide the file names, with the
complete file path, to the function `readSAS`,

```{r}
library(OzoneExposure)
ozone_files = c(
"~/Downloads/z9312_epa.sas7bdat",
"~/Downloads/z9312_kim.sas7bdat",
"~/Downloads/z9312_ucd.sas7bdat")

ozone_data = readSAS(ozone_files)
```

The `readSAS` function will take multiple file names concatonated
together, allowing us to read one or all of the data into R at once.

### Subsetting and Formatting Data

The data are formatted for SAS, with multiple rows per
individual/experiment representing different dFEV1 measument times,
but with the intervals for changes in O3 or exercize represented by
multiple columns. To fit the models to these data, we need to change
the format. The function `mungeSAS`
will re-format the data, but we should first do any subsetting of
individuals or experiments first.

For example, lets say we are only interested in the data from the
"ARO" experiment,

```{r}
aro_data = subset(ozone_data, STUDY == "ARO")

```

We can now re-format these data,

```{r}
aro_model_data = mungeSAS(aro_data)
```

The reformatted data are in a list, with multiple matrices
representing the different variables (e.g., O3, Ve, or BSA
levels). 

### Fitting the Models

The list object created by `mungeSAS` can directly be fed into a
model, for example the McDonnell model;

```{r}
result = fit_McDonnel(aro_model_data, n_optim = 10)
min(result$aic)
```

or the Schelegle model,

```{r results = "hide"}
result2 = fit_Schelegle(aro_model_data, n_interval = 5)
```

```{r}
result2[which.min(result2$aic),]
```

We can then explore which combination of parameters give us the lowest
AIC values,

```{r}
sapply(results
